---
title: "NMDS analysis of transcriptomic data"
author: "Maggi Brisbin"
date: "2023-10-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## packages

```{r loadpackages, warning=FALSE, message=FALSE}
library(tidyr)
library(dplyr)
library(ggplot2)
library(vegan)
library(DESeq2)
library(phyloseq)
```

## data
import data and format for deseq2:
```{r}
exp <- read.csv("transcript_df.csv")
row.names(exp) <- exp$orf_id
exp <- exp[,-1]
exp<- exp[-c(1,2),]

meta <- read.csv("transcript_df_anno.csv")
meta$X <- paste("X", meta$X, sep="")
row.names(meta) <- meta$X
meta <- meta[,-1]
meta$b12 <- factor(meta$b12)
```


## deseq2 variance stabilization
convert data to deseq2 experiment object:
```{r, warning=FALSE}
dds<- DESeqDataSetFromMatrix(countData = exp,
                              colData = meta,
                              design = ~ iron)
dds
```
variance stabilize:
```{r}
vsd <- vst(dds, blind=TRUE)
```



## NMDS
convert variance stabilization results to a dataframe and a matrix:
```{r}
vsd_df <- data.frame((assay(vsd)))
vsd_mat <- data.matrix(vsd_df)
dim(vsd_mat) #check data dimensions
```
`49068 contigs, 36 samples`

convert data matrix to phyloseq object:
```{r}
VSD <- otu_table(vsd_mat, taxa_are_rows = TRUE)
META <- sample_data(meta)
PS <- phyloseq(VSD, META)
```

ordination of euclidean distances between samples:
```{r}
ord <- ordinate(PS, "NMDS", "euclidean", k=4, trymax=50)
```
Plot NMDS ordination results:
```{r}
p<-plot_ordination(PS, ord, type="samples")+theme_bw()  +  theme(text = element_text(size=14)) +  geom_hline(yintercept = 0, linetype = "dashed", color = "lightgrey") +  geom_vline(xintercept = 0, linetype = "dashed", color = "lightgrey") + theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank()) +geom_point(aes(fill=b12, shape = iron), size =3)  + scale_shape_manual(values= c(21,24,22)) +scale_fill_manual(values=c("#333ED4", "#A1D5DE", "#A0D636", "#EEDE04", "#FFA52C", "#FD0100"))  +guides(fill = guide_legend(override.aes=list(shape=21)))

p
```

`ggsave("NMDS_newcolors.pdf", width = 6, height = 4)`

## PERMANOVA
perform PERMANOVA on euclidean distances between samples:
```{r}
set.seed(1)

vdist = vegdist(t(vsd_mat), method = "euclidean")

adonis2(vdist ~ iron+b12, data = meta)
```
## ANOSIM

```{r}
ano<- with(meta, anosim(vdist, condition))
summary(ano)
```

